Cheminformatics: The Computer Science of Chemical Discovery

### 2011-11-10 rebuttal.tex

\comment{p1, c1, l32: I think the phrase miracle molecule could be
  more usefully replaced with new drugs or new small molecule
  therapeutics.}
\response{The text has been updated to use the term ``therapeutic
  molecule''}

\comment{p2, c2, l57: it might be worthwhile here clarifying in the text
precisely the properties that require satisfaction to deliver a small
molecule therapeutic, not ?promising?. A drug must be safe and
efficacious ultimately; perhaps this should be mentioned first
followed by the typical pitfalls and how they are assessed?}
\response{The text has been updated to provide some examples of
  properties that would characterize a therapeutically useful small molecule}

\comment{P3, c1, l41: the authors need to cover the Morgan algorithm
  (published in 1965) here and explain the canonicalization
  process. The issues mentioned in different canonicalization
  implementations providing different SMILES strings is also a
  challenge with InChI codes with different softwares giving different
  representation. Therefore, this is still not a solved problem as
  suggested here. Could the authors clarify this in the text?}
\response{We have updated the text to reference the Morgan
  algorithm. Regarding the issue of InChI codes, we believe that this
  is not the case. Currently, there is only one implementation of the
  InChI algorithm and even if alternative implementations were to be
  developed, the InChI specification is publically available. As a
  result, there should be no differing InChI representations for a
  given input structure.}

\comment{P3, c2: structure-based fingerprints referred to here sound
  like Daylight-style fingerprints to this reviewer. Is this the case?
  I would emphasise the distinction of structure-key and hash-key
  fingerprints. My understanding is that Daylight-style fingerprints
  enumerate paths of 7 in length, not 8, could the authors provide a
  reference for this?}
\response{The section on fingerprints has been rewritten to be more
  clearer as well as distinguish between the two braod classes of
  fingerpritns. We have also mentioned the use of fingerprints in
  similarity searchs and the Tanimoto coeffcient}

### 2011-11-10 paper.tex
Added some more text on fingeprints

InChI or other identity-mapping algorithms allow for exact searching
(within the limitations of the algorithms). However, two other
practically relevant algorithms for scientific discovery based on
chemical databases are \emph{substructure searching} and
\emph{similarity searching}, which are required to generalize from the
data points contained in a particular database to other, related
molecules. In substructure searching, the database is searched for a  
specified wholly contained part of the structures in the database,
while in similarity searching, the database is searched for structures
similar (in some property space) to a provided search, or query,
structure. Chemical search packages are often implemented and
optimized for a given database technology; for example, the OrChem
package is an open source chemical search package for the Oracle
database \cite{rijnbeek2009}.

Graph substructure matching is a variant of \emph{graph isomorphism
  detection}, which is widely believed to be computationally
intractable~\cite{cordella2001}. To execute a graph isomorphism search
across a full chemical database of thousands or even millions of
structures is infeasible \cite{Weininger:2011ly}. For that reason,
heuristics must be used to drastically filter the candidate search
space. Structure-based \emph{fingerprints} are commonly used for this
purpose. Fingerprints encode characteristic features of a given
chemical structure in a Boolean array, or bitmap, and they can be
implemented in a variety of ways. Common fingerprints are created by
an algorithm which generates bitmap patterns for each atom in the
structure; then each atom and its nearest neighbors, including the
bonds between them; then each group of atoms connected by paths of
increasing lengths up to the maximum implemented for the algorithm
(commonly around eight). In this way, characteristic parts of molecule
are captured and the bitmap patterns are added together into the final
fingerprint. If all bits in a query fingerprint are also present in
the target fingerprint of a stored database structure, only then is
the structure subjected to the computationally expensive subgraph
matching algorithm. Bit operations are very fast and independent of
the number of atoms in a structure due to the fixed length of the
fingerprint.

@@

Enumerating molecules is a combinatorial problem that has fascinated
chemists, computer scientists and mathematicians alike, for more than
a century. It is one important risk-reduction element in still being
able to encode closely-related molecules and for keeping the space
complexity of databases under (algorithmic) control. Indeed, while
trying to solve the problem of counting the isomers of paraffin
structures or counting substituted aromatic compounds, fundamental
principles in graph theory and combinatorics were developed by Cayley,
Polya and others. Even terms that are widely-used nowadays, like
``graph'' and ``tree'', were originally coined in a chemistry context
(Chapters 1 and 8 in \cite{faulon2010}). More than four decades ago,
Joshua Lederberg, Carl Djerassi, Edward Feigenbaum and others from
Stanford University started developing codes to explicitly enumerate
molecules. DENDRAL is generally quoted in artificial intelligence
textbooks as a pioneer project that produced the first expert
system~\cite{DENDRAL}. However, even before that

>>>>

InChI or other identity-mapping algorithms allow for exact searching. However, two other
practically relevant algorithms for scientific discovery based on
chemical databases are \emph{substructure searching} and
\emph{similarity searching}, which are required to generalize from the
data points contained in a particular database to other, related
molecules. In substructure searching, the database is searched for a  
specified wholly contained part of the structures in the database,
while in similarity searching, the database is searched for structures
similar (in some property space) to a provided search, or query,
structure. Chemical search packages are often implemented and
optimized for a given database technology; for example, the OrChem
package is an open source chemical search package for the Oracle
database \cite{rijnbeek2009}.

Graph substructure matching is a variant of \emph{graph isomorphism},
which is widely believed to be computationally
intractable~\cite{cordella2001}. To execute a graph isomorphism search
across a full chemical database of thousands or even millions of
structures is infeasible \cite{Weininger:2011ly}. Speedups can be
obtained via the use of heuristics such as structural
\emph{fingerprints}. Fingerprints encode characteristic features of a
given chemical structure, usually in a fixed-length
bitmap. Fingerprints fall broadly into two categories: structure keys
and hashed keys. In the former, each bit position corresponds to a
distinct substructure such as an alcohol or keto group. Examples
include MACCS and PubChem keys. The latter class are designed such
that substructural patterns are represented as strings and then hashed
to a random bit position. As a result, a given position can encode
multiple substructures. The advantage of such fingerprints is that
they can cover an arbitrarily large collection of substructures such
as paths of length $N$ or circular environments and so on. Examples
include ECFPs and Daylight fingerprints.

Given a binary fingerprint, we can first ``pre-screen'' a database, to
ignore molecules that cannot possibly match the query. This is
achieved by requiring that all bits in a query fingerprint must also
be present in the target fingerprint. Since the target fingerprints
are pre-computed, comparing binary strings can be performed extremely
rapidly on modern hardware. As a result, we apply the actual
isomoprhism test only on those molecules that pass the
screen. Fingerprints can also be used to rapidly search databases for
similar molecules, in which case a similarity metric such as the
Tanimoto coefficient is used to compare the query and target
fingerprints. While fast, heuristics have been developed
\cite{Swamidass:2007ve} to further
speed up similarity searches.

@@

Enumerating molecules is a combinatorial problem that has fascinated
chemists, computer scientists and mathematicians alike, for more than
a century. It is one important risk-reduction element in still being
able to encode closely-related molecules and for keeping the space
complexity of databases under (algorithmic) control. Indeed, while
trying to solve the problem of counting the isomers of paraffin
structures or counting substituted aromatic compounds, fundamental
principles in graph theory and combinatorics were developed by Cayley,
Polya and others. Even terms that are widely-used nowadays, like
``graph'' and ``tree'', were originally coined in a chemistry context
(Chapters 1 and 8 in \cite{faulon2010}). More than four decades ago,
Joshua Lederberg, Carl Djerassi, Edward Feigenbaum and others from
Stanford University started developing codes to explicitly enumerate
molecules. DENDRAL is generally quoted in artificial intelligence
textbooks as a pioneer project that produced the first expert
system~\cite{DENDRAL}.
